Functional Pearl --- Programming with \\ Negative, Fractional, Square Root, and Imaginary Types

### 2012-02-20 dual.tex
added stuff on imaginary and square root types

\title{Functional Pearl \\ Programming with Negative, Fractional, and Imaginary Types} 
\authorinfo{Roshan P. James}
           {Indiana University}
           {rpjames@indiana.edu}
\authorinfo{Amr Sabry}
           {Indiana University}
           {sabry@indiana.edu}
\maketitle

\begin{abstract}
  Every functional programmer knows about sum and product types, $a+b$ and
  $a*b$ respectively. Negative, fractional, and imaginary types, $a-b$,
  $a/b$, and $a+ib$ respectively, are much less known and their computational
  interpretation is unfamiliar and often complicated. We show that in a
  programming model in which information is preserved (such as the model
  introduced in our recent paper on \emph{Information Effects}), negative,
  fractional, and imaginary types have particularly intuitive and natural
  computational interpretations. Intuitively, values of negative types are
  values that flow ``backwards'' to satisfy demands, values of fractional
  types are values that represent first-class constraints, and values of
  imaginary types are values that capture the essence of recursive datatype
  declarations. The combination of negative, fractional, and imaginary types
  allow a programmer to express a wide range of programming idioms, including
  higher-order functions, delimited continuations, speculative computation,
  logic programming, and more.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Imaginary Types: Intuition}

Imaginary types have also appeared in the literature: we relegate the
connections to Sec.~\ref{related} and proceed with a simple explanation. We
have so far extended the set of types to be the rational numbers. Now we will
push this and extend the set of types to algebraic numbers. In other words,
we will allow datatypes defined by arbitrary polynomials and allow the roots
of such polynomials to be types. Consider the type of binary trees:
\begin{verbatim}
data B = 1 + B * B
\end{verbatim}
In conventional settings, this declaration introduces an isomorphism between
$b$ and $1+b^2$ (whose two sides are commonly known as {{fold}} and
{{unfold}}). In our setting, this induces an isomorphism between $0$ and $b^2
- b + 1$. The latter polynomial has roots $(1 \pm i\sqrt{3}) / 2$ and hence
we have two isomorphisms between $b - (1 + i\sqrt{3})$ and $0$ and also
between $b - (1 - i\sqrt{3})$ and $0$.

Now say I want to produce a value of type $b$. We can use the isomorphisms
above to break this task in many ways. Essentially exploit that $b$ is
isomorphic to $b^7$, let the various program components produce the various
parts of $b^7$ and merge them.

Consider some simple examples first.

\item $b=1+1$. Boring.

\item $b=b+1$. That introduces the natural numbers. No solution for this
  polynomial over the algebraic numbers. We must extend the numbers with
  $\omega$ to get a solution. We reject this in this paper and prefer to
  stick to algebraic numbers. The advantage is that all the isomorphisms are
  valid numerically. With the above type we could subtract $b$ from both
  sides to show that $0=1$ which is nonsense. We can however have infinite
  types as long as they have algebraic solutions.

\item $b^2=2$ or $b = \pm \sqrt{2}$. We have introduced the square root of a
  boolean! If we had superpositions, we could then write a function that
  performs the square root of negation in the sense that applying it twice
  would be equivalent to boolean negation.

>>>>

\title{Functional Pearl --- Programming with \\ 
  Negative, Fractional, Square Root, and Imaginary Types} 
\authorinfo{Roshan P. James}
           {Indiana University}
           {rpjames@indiana.edu}
\authorinfo{Amr Sabry}
           {Indiana University}
           {sabry@indiana.edu}
\maketitle

\begin{abstract}
  Every functional programmer knows about sum and product types, $a+b$ and
  $a*b$ respectively. Negative, fractional, square root, and imaginary types,
  $a-b$, $a/b$, $\sqrt{a}$, and $a+ib$ respectively, are much less known and
  their computational interpretation is unfamiliar and often complicated. We
  show that in a programming model in which information is preserved (such as
  the model introduced in our recent paper on \emph{Information Effects}),
  these types have particularly intuitive and natural computational
  interpretations. Intuitively, values of negative types are values that flow
  ``backwards'' to satisfy demands, values of fractional types are values
  that represent first-class ``structural'' constraints, and values of square
  root and imaginary types are values that can be related by arbitrary
  algebraic constraints. The combination of these negative, fractional,
  square root, and imaginary types enable greater flexibility in programming
  by breaking complicated invariants into local ones that can be
  independently satisfied by a subcomputation. Furthermore, they allow a
  programmer to express a wide range of programming idioms, including
  higher-order functions, delimited continuations, speculative computation,
  logic programming, and more.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Algebraic Types: Intuition}

Square root and imaginary types have also appeared in the literature: we
relegate the connections to Sec.~\ref{related} and proceed with a simple
explanation. We have so far extended the set of types to be the rational
numbers. Now we will push this and extend the set of types to algebraic
numbers. In other words, we will allow datatypes defined by arbitrary
polynomials and allow the roots of such polynomials to be types.

Consider first an example in which we want to compute with the sides of a
rectangle whose area is 91 and whose length is longer than its width by 6
units. One can solve the quadratic equation to determine that the sides are 7
and 13 and proceed. This however prematuraly forces us to globally solve the
constraint. Instead we can let the two sides of the rectangle be $x$ and
$x+6$ and use the following equation to capture the desired constraint:
\[
x^2 + 6x - 91 = 0
\]
The equation introduces an isomorphism between the type $x^2 + 6x - 91$ and
the type $0$. We can now proceed to compute with the unknown $x$, being
assured that in a closed program, our computation will eventually be
consistent with the solution of the algebraic equation.

Previously the most famous example of a similar nature is the puzzling
isomorphism that one can establish between seven binary trees and one.
A binary tree is defined by the datatype
\[
x = 1 + x * x 
\]
which can be rearranged to the polynomial equation $x^2 - x + 1 = 0$. By
algebraic manipulation we can reason as follows:
\[\begin{array}{rclclclcl}
x^3 &=& x^2 x &=& (x-1) x &=& x^2 - x &=& -1 \\
x^6 &=& 1 \\
x^7 &=& x^6 x &=& x
\end{array}\]
Fiore poses the question of why such algebraic manipulation would make sense
type theoretically but states that even though some of the intermediate steps
make no sense, the final equivalence is valid and can be used to actually
construct an isomorphism between $x^7$ the type of seven binary trees 
and $x$ the type of binary trees.

Discussion of possible polynomials:
\begin{itemize}
\item $b=1+1$. Boring.
\item $b=b+1$. That introduces the natural numbers. No solution for this
  polynomial over the algebraic numbers. We must extend the numbers with
  $\omega$ to get a solution. We reject this in this paper and prefer to
  stick to algebraic numbers. The advantage is that all the isomorphisms are
  valid numerically. With the above type we could subtract $b$ from both
  sides to show that $0=1$ which is nonsense. We can however have infinite
  types as long as they have algebraic solutions.
\item $b^2=2$ or $b = \pm \sqrt{2}$. We have introduced the square root of a
  boolean! If we had superpositions, we could then write a function that
  performs the square root of negation in the sense that applying it twice
  would be equivalent to boolean negation.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Summary}

To summarize negative, fractional, square root, and imaginary types all make
sense. What they help you accomplish as a programmer is to disassociate
global invariants into local ones that can be satisfied independently by
subcomputations with no synchronization or communication. A computation
producing something of type $a/b$ does not need to concern itself with who is
going to supply the missing $b$: it just does its part. Conversely faced with
a complicated task, a computation might decide to break it into pieces and
demand these pieces using negative types.

It is no surprise that these types are closely related to quantum mechanics
and that they give us the feel that this is how nature computes. This is
speculation however.

In any case, in a framework where information can be copied and deleted, none
of this makes much sense. It is critical that these constraints and demands
can neither be duplicated nor erased.  This gives us the maximum
``parallelism'' possible.
